---
title: "Yellen dashboard"
author: "Mitsuo Shiota"
date: "2019-03-19"
output: 
  github_document:
    toc: TRUE
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Updated: `r Sys.Date()`

## Summary

[Dashboard in pdf](output/Yellen.pdf)

This is a pilot project to utilize R language and tidyquant package for economic analysis.

I read the bloomberg article, ["Yellen’s Labor Market Dashboard"](https://www.bloomberg.com/graphics/2015-yellens-labor-market-dashboard/), and found 12 time-series data, central bankers are supposed to watch. I decided to replicate this dashboard. I added PCE, excl food and energy, to see price stability. Update: I also added prime age employment rate, as it is said to be more closely related to wage than headline (un)employment rate.

## Libraries

I use tq_get function from tidyquant package to get tidy data from [Federal Reserve Bank of St. Louis (FRED)](https://fred.stlouisfed.org/). And some of my self-made functions depend on tidyquant package.

I use GET function from httr package to get an excel file from [Federal Reserve Bank of Atlanta (Wage Growth Tracker)](https://www.frbatlanta.org/chcs/wage-growth-tracker.aspx), and readxl package to read excel data. 

```{r libraries, message=FALSE}
library(tidyquant)
library(tidyverse)
library(httr)
library(readxl)

```

## Self-made tqr package

tidyquant package assumes tidy data, basically a data frame of long format, of 3 columns: "date", "symbol" and "price." These column names come from stock price analysis. I use self-made [tqr package](https://github.com/mitsuoxv/tqr) to transform values in price column.

```{r tqr, message=FALSE}
library(tsibble)
library(tqr)

```

## Get data from FRED

At [FRED](https://fred.stlouisfed.org/), I searched data, and got symbols from the parentheses in the title. For example, if the title is "Japan / U.S. Foreign Exchange Rate (AEXJPUS) ", "AEXJPUS" is the symbol.

```{r FRED_symbols}
yellen_labor_mkt_symbols <- c(
  "UNRATE", # Unemployment rate
  "U6RATE", # U-6, underemployment rate
  "LNS13025703", # Long-term unemployed share
  "PAYEMS", # Non-farm payrolls, total
  "JTSJOR", # Job openings rate
  "JTSLDR", # Layoffs/discharged rate
  "JTSQUR", # Quits rate
  "JTSHIR", # Hires rate
  "CEU0500000003", # Average hourly earnings, total private
  "ECIALLCIV", # Employment cost index, all civilians, quaterly
  "CIVPART", # Labor force participation rate
  "LREM25TTUSM156S", # Employment Rate: Aged 25-54: All Persons
  "PCEPILFE" # PCE excluding food and energy
  ) 

```

I set the first date to get data as START. I also set XLIM, x axis limits, to plot later.

```{r set_START}
START = "2006-01-01"

XLIM <- c(as.Date("2008-01-01"), as.Date("2025-12-01"))

```

Now I can use tq_get function from tidyquant package to download data from FRED.

```{r get_FRED}
labor_mkt_all <- yellen_labor_mkt_symbols %>% 
  tq_get(get = "economic.data", from = START)

```

## Transform data

Employment cost index is the only quarterly data, and all the others are monthly data. I transform Employment cost index to year-over-year growth rates, and all the others to tsibble (tbl_ts class).

```{r transform_0}
# transform Employment cost index, quarterly data to YoY
eci <- labor_mkt_all %>% 
  filter(symbol == "ECIALLCIV") %>% 
  mutate(date = yearquarter(date)) %>% 
  as_tsibble(key = "symbol", index = "date") %>% 
  tq_gr(n = 4) %>% 
  mutate(date = as.Date(date))

# Others are monthly data, transformed to tsibble
labor_mkt_m <- labor_mkt_all %>% 
  filter(symbol != "ECIALLCIV") %>% 
  mutate(date = yearmonth(date)) %>% 
  as_tsibble(key = "symbol", index = "date")

```

Non-farm payrolls requires the most complicated transformation. I have to take differences from the prior month, and make them 3 month moving average. Fortunately pipe operator helps me.

```{r transform_1}
# transform Non-farm payrolls, total to differences, 3 month moving average
payems <- labor_mkt_m %>% 
  filter(symbol == "PAYEMS") %>% 
  tq_diff() %>% 
  tq_ma(n = 3)

```

I transform PCE to year-over-year growth rates, and average hourly earnings to year-over-year growth rates and 3 month moving average.

```{r transform_2}
# transform PCE to growth rates, YoY
pce <- labor_mkt_m %>% 
  filter(symbol == "PCEPILFE") %>% 
  tq_gr(n = 12)

# transform Average hourly earnings to YoY, 3 ma
ceu <- labor_mkt_m %>% 
  filter(symbol == "CEU0500000003") %>% 
  tq_gr(n = 12) %>% 
  tq_ma(n = 3)


```

## Get data from Atlanta Fed Wage Growth Trucker

There is one datum I can't get from [FRED](https://fred.stlouisfed.org/), that is [Atlanta Fed Wage Tracker](https://www.frbatlanta.org/chcs/wage-growth-tracker.aspx). Unfortunately I could not find API to get data. So, initially I manually downloaded an excel file,  transformed it into a csv file, and used read_csv to read.

Later, I found ["Read Excel file from a URL using the readxl package"](https://stackoverflow.com/questions/41368628/read-excel-file-from-a-url-using-the-readxl-package) in StackOverflow. Now, I automate the reading process, though it is awkward. If Atlanta Fed changes something, this part will fail.

```{r Atlanta, results="hide", message=FALSE}
url <- "https://www.frbatlanta.org/-/media/documents/datafiles/chcs/wage-growth-tracker/wage-growth-data.xlsx"

GET(url, write_disk(tf <- tempfile(fileext = ".xlsx")))

res <- read_excel(tf, col_names = FALSE, na = c(".", ""), skip = 2)

names(res)[1:2] <- c("date", "price")
res$date <- as.Date(res$date)

wage_tracker <- res %>% 
  select(date, price)
  
wage_tracker$symbol <- "WAGETR"
wage_tracker <- wage_tracker %>% 
  filter(date >= START)

```

## Combine data

I combine transformed data with not-transformed data, and get tidy data. As `bind_rows` function drops index necessary to tbl_ts class, I have to transform tsibble to tibble.

```{r combine}
labor_mkt <- labor_mkt_m %>% 
  filter(!(symbol %in% c("PAYEMS", "PCEPILFE", "CEU0500000003"))) %>% 
  bind_rows(payems, pce, ceu, eci, wage_tracker) %>% 
  as_tibble()

```

## Prepare for the plot

I change class of symbol from character to factor in the appropriate order for the plot. 

```{r symbol_factor}
labor_mkt$symbol <- factor(labor_mkt$symbol,
                           levels = c("UNRATE", "LNS13025703", "U6RATE", "CIVPART",
                                      "JTSJOR", "JTSLDR", "JTSQUR", "JTSHIR",
                                      "PAYEMS", "CEU0500000003", "ECIALLCIV", "WAGETR",
                                      "LREM25TTUSM156S", "PCEPILFE"))


```

I recode symbol to make them more readable in the plot.

```{r symol_recode}
labor_mkt <- labor_mkt %>% 
  mutate(
         symbol = fct_recode(symbol,
                             "Unemployment rate" = "UNRATE",
                             "Long-term unemployed share" = "LNS13025703",
                             "U-6 underemployment rate" = "U6RATE",
                             "Labor force participation rate" = "CIVPART",
                             "Job openings rate" = "JTSJOR",
                             "Layoffs/discharged rate" = "JTSLDR",
                             "Quits rate" = "JTSQUR",
                             "Hires rate" = "JTSHIR",
                             "Non-farm payroll increase, 3 mo avg" = "PAYEMS",
                             "Avg hourly earnings, YoY, 3 mo avg" = "CEU0500000003",
                             "Employment cost index, YoY, quarterly" = "ECIALLCIV",
                             "Atlanta Fed wage tracker, YoY, 3 mo avg" = "WAGETR",
                             "Employment Rate: Aged 25-54" = "LREM25TTUSM156S",
                             "PCE excluding food and energy, YoY" = "PCEPILFE"
                             )
           )

```


## Dashboard

OK. Let us plot. I add the labels at the latest. Some of them are put ugly, but useful enough.  

```{r plot,fig.width=8,  fig.height=16}
labor_mkt %>% 
  group_by(symbol) %>% 
  mutate(
    label = if_else(date == max(date), 
                    paste(lubridate::month(date, label = TRUE, abbr = TRUE),
                          as.character(round(price, 2))
                          , sep = " "),
                    NA_character_)
  ) %>% 
  ungroup() %>% 
  ggplot(aes(x = date, y = price)) + 
  geom_line() +
  facet_wrap(~ symbol, ncol = 2, scales = "free_y") +
  geom_text(aes(label = label), na.rm = TRUE,
            hjust = 1, vjust = 0) +
  coord_cartesian(xlim = XLIM) +
  labs(
    title = "Yellen's US labor market dashboard",
    x = "",
    y = "")

```

Save plot in output/Yellen.pdf.

```{r plot_save}
ggsave(filename = "output/Yellen.pdf",
       width = 8, height = 16, units = "in", dpi = 300)

```

EOL